---
title: "Getting Started and Setting Up `freesurfer` in R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started and Setting Up `freesurfer` in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ref.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  eval = freesurfer::have_fs(),
  prompt = FALSE,
  message = FALSE,
  warning = FALSE
)
```

Freesurfer is a widely-used software package for processing and analyzing anatomical neuroimaging data [@fischl2012freesurfer].
Developed by the Laboratory for Computational Neuroimaging at the Athinoula A. Martinos Center for Biomedical Imaging, it offers a suite of open-source, command-line tools.
These tools handle essential image processing tasks such as:

* Brain extraction or skull-stripping, which removes non-brain tissues from images [@segonne2004hybrid].
* Bias-field correction, which corrects for intensity non-uniformities in MRI scans [@sled_nonparametric_1998].
* Segmentation of brain structures, allowing for the identification and measurement of different regions [@fischl2002whole; @fischl2004sequence].
* Image registration, which aligns different brain images to a common space [@fischl1999high; @reuter2010highly].

Beyond these individual functions, Freesurfer also provides fully-automated pipelines that streamline complex processing workflows.

While R offers several powerful packages for image data, such as `AnalyzeFMRI` [@bordier_temporal_2011] and `fmri` [@tabelow_statistical_2011] for functional MRI analysis and spatial smoothing, `RNiftyReg` [@modat_rniftyreg:_2013] for image registration, and `dpmixsim` [@dpmixsim] and `mritc` [@mritc] for image clustering and segmentation (see the [Medical Imaging CRAN task view](http://cran.r-project.org/web/views/MedicalImaging.html) for more), the neuroimaging community has developed even more specialized tools that might perform better on specific datasets or offer more comprehensive information.
Freesurfer, for instance, includes methods not currently implemented in R, such as surface-based registration and completely automated image segmentation pipelines.

The `ANTsR` package (available on [GitHub](https://github.com/stnava/ANTsR)), an unpublished R package, has implemented additional image analysis functionality, but it doesn't encompass everything Freesurfer offers.
Having multiple options for image processing directly within R empowers users to compare different methods and provides the flexibility to combine various packages to create a robust data processing pipeline.

This `freesurfer` package serves as an interface, allowing R users to leverage Freesurfer's state-of-the-art anatomical processing capabilities, along with a suite of tools that simplify analyzing Freesurfer's output.
Our goal is to enable R users to perform complete anatomical imaging analyses without needing to learn Freesurfer's specific command-line syntax, keeping both the image processing and analysis seamlessly integrated within R.

## Setting Up Your R Environment for `freesurfer`

To use the `freesurfer` R package, you'll first need a working installation of Freesurfer itself.
You can download and install Freesurfer from its [official website](http://freesurfer.net/fswiki/DownloadAndInstall).

The `freesurfer` R package relies on successfully detecting your Freesurfer installation.
The core of this detection process is handled by a series of functions `get_fs_home()`, which attempts to locate your `FREESURFER_HOME` directory.

The detection process for `FREESURFER_HOME` follows a specific order:

1. It looks for the R option `getOption("freesurfer.home")`.   
1. If that is not found, `freesurfer` checks your system's environment variables for `FREESURFER_HOME`.   
1.  As a last resort, it tries common default installation paths, such as: 
    - `/usr/local/freesurfer`  
    - `/Applications/freesurfer`  
    - `/usr/freesurfer`
    - `/usr/bin/freesurfer`  

As a first step, use the `fs_sitrep()` function to get detailed information about the setup between R and Freesurfer on your system.

```{r fs_sitrep}
library(freesurfer)

fs_sitrep()
```

This function provides useful information about which issues may be present regarding the integration of R and Freesurfer on your system.
Ideally all checks should pass with a green tick.
Any red ticks or warnings should be addressed to ensure smooth use of this package.
As you can see, there are several things to make sure are in order.
As a minimun, you should have the  `freesurfer.home` option or `FREESURFER_HOME` variable set.
If you see a red tick for `freesurfer.home` or `FREESURFER_HOME`, it means that the package cannot find your Freesurfer installation.
However, as long as this is set and has a green tick, the remaining options are likely set to sensible defaults.

### Setting Up Freesurfer Options

When you're running R from a shell environment, `freesurfer` will automatically use the `FREESURFER_HOME` environment variable (which is typically set during Freesurfer installation) as the path to your Freesurfer directory.
However, if you're using R through a graphical user interface (GUI) like RStudio, environment variables and paths might not be explicitly exported, meaning `FREESURFER_HOME` might not be set.
In such cases, `freesurfer` will try the default directories for macOS and Linux.
If your Freesurfer installation isn't in a standard location, you can manually specify its path using `options(freesurfer.home="/path/to/freesurfer")`.
Please note that Freesurfer is only available on Windows via a virtual machine.

While R may detect your system variables, as mentioned above, any `options` set in either your project or user `.Rprofile`, or in the script you are running, will take precedence over the system settings.
This makes is possible for you to set script or project specific options and get more control over how Freesurfer is used in your analyses.

To set these options in your `.Rprofile` file, you can open these with functions like `usethis::edit_r_profile("user")` and `usethis::edit_r_profile("project)`.
We highly recommend using the project `.Rprofile` if you are in a shared environment, have a project several people access, and have a common installation of Freesurfer to access.
On your personal computer, its likely you'd rather set the variables in your user `.Rprofile`.
Possible options to set are:

```r
options(
    freesurfer.home = "/path/to/freesurfer",
    freesurfer.subj_dir = "/path/to/freesurfer/subjects",
    freesurfer.license = "/path/to/freesurfer/.license",
    freesurfer.output_type = "hdr",
    freesurfer.sh = "/path/to/freesurfer/FreeSurferEnv.sh",
    freesurfer.mni_dir = "/Applications/freesurfer/mni/Library/Perl/Updates/5.12.3",
    freesurfer.verbose = TRUE
)

# or using environment variables
Sys.setenv(
    FREESURFER_HOME = "/path/to/freesurfer",
    FREESURFER_SUBJ_DIR = "/path/to/freesurfer/subjects",
    FS_LICENSE = "/path/to/freesurfer/.license",
    FSF_OUTPUT_FORMAT = "hdr",
    FREESURFER_SH = "/path/to/freesurfer/FreeSurferEnv.sh",
    MNI_DIR = "/Applications/freesurfer/mni/Library/Perl/Updates/5.12.3",
    FREESURFER_VERBOSE = TRUE
)
```

But if you are more comfortable using environment variables, you can also both have user and project specific `.Renviron`, just like `.Rprofile`. 
You can open these with functions like `usethis::edit_r_environ("user")` and `usethis::edit_r_environ("project)`

Once the options are set, restart your R session, and try another `fs_sitrep()` which should now use the options you have provided.
The `freesurfer.verbose` option is particularly useful for debugging, as it will print out additional information about the Freesurfer commands being executed.


### Check if Freesurfer is installed

The `have_fs` function is a handy tool that checks if you have a Freesurfer installation.
It returns a logical value (TRUE or FALSE), which is very useful for conditional statements in your code.
You can also ask it to check for the presence of a license file by specifying `check_license = TRUE`.

```{r have_fs_example}
have_fs()

if (have_fs()) {
  print("Freesurfer is installed! Huzzah!")
}
```

Be aware that the `have_fs()` function is simple in that it check if the `fs_dir()` actually exists or not.
It does not make any validation of whether this Freesurfer installation and R are set up correctly to work together.
So make sure your `fs_sitrep()` looks good, first and foremost, then `have_fs()` is a convenient tool for scripting logic.

## Understanding Freesurfer Output Formats

Freesurfer commands often default to a specific output format for image files, which is controlled by the option `freesurfer.output_type` or `FSF_OUTPUT_FORMAT` environment variable.
You can check this setting using `get_fs_output()`.
If it's not explicitly set, it typically defaults to `nii.gz`.
You can also retrieve the corresponding file extension with `fs_imgext()`.

```{r fs_output_example}
get_fs_output()
fs_imgext()
```

### Understanding how `get_fs()` fascilitates Freesurfer calls

The `get_fs()` function is a crucial utility in this R package for interacting with FreeSurfer. 
Its primary purpose is to generate a shell command string that correctly configures your environment to run FreeSurfer commands, ensuring that the FreeSurfer executables, license, and other necessary settings are properly recognized by your system.

#### Why is `get_fs()` important?

As described above, FreeSurfer relies on specific environment variables and setup scripts to function correctly.
Without these, your R session wouldn't know where to find the FreeSurfer programs (like `recon-all` or `mri_convert`), or how to handle their output. 
While the `fs_sitrep()` will help you get setup up correctly, its the `get_fs()` that powers the Freesurfer-R integration after this.
Based on the settings you have provided (or default ones if you have not), `get_fs()` will create a string that will prefix any Freesurfer command that is called from R.
Depending on your system setup, and the complexity of the Freesurfer function called, this string is a vital component to running the Freesurfer commands from R and capturing their results.


#### How it works

`get_fs()` follows a logical hierarchy to determine and set up your FreeSurfer environment:

1.  **Finding `FREESURFER_HOME`** as described above

1.  **License Check**:

      * Once `FREESURFER_HOME` is identified, `get_fs()` checks for the presence of a FreeSurfer license file (either `license.txt` or `.license`) within that directory.
      * If the license file is not found, a warning will be issued, as FreeSurfer may not run correctly without a valid license.

1.  **Handling Binary Application Directory (`bin_app`)**:

      * The `bin_app` argument allows you to specify a subdirectory within `FREESURFER_HOME` where the main executables are located. Common values are:
          * `"bin"`: This is the most typical location for FreeSurfer's core binaries.
          * `"mni/bin"`: If you need to use FreeSurfer's MNI (Montreal Neurological Institute) related tools, this option will ensure that the necessary MNI Perl startup scripts are initialized, which is crucial for those functions to work.
          * `""`: This option is for cases where the executables are directly in the `FREESURFER_HOME` directory, without a subdirectory.

1.  **Sourcing `FreeSurferEnv.sh`**:

      * FreeSurfer often includes a shell script called `FreeSurferEnv.sh` in its installation directory. This script performs additional, more detailed environment configurations.

1.  **Setting Output Format**:

      * The function also sets the `FSF_OUTPUT_FORMAT` environment variable, which dictates the default image format for FreeSurfer's output (e.g., `.nii.gz` for compressed NIfTI files).

1.  **Generating the Command String**:

      * Finally, `get_fs()` consolidates all these steps into a single bash command string. This string includes `export` commands for environment variables and commands to source necessary setup scripts, followed by the path to the FreeSurfer binary directory. This complete command string is what `fs_cmd()` (and potentially other functions in your package) will execute in the background to run FreeSurfer tools.

### Example Usage

In most cases, you will never directly need to run this command.
But it is good to know of it, how it works and what it's output is.

```{r}
# Generate a shell command to set up FreeSurfer with the default `bin` directory
get_fs(bin_app = "bin")

# Generate a shell command to include MNI environment setup
# This is important if you plan to use FreeSurfer's MNI tools.
get_fs(bin_app = "mni/bin")
```

By using `get_fs()`, you ensure that your R interactions with FreeSurfer are robust and correctly configured, abstracting away the complexities of shell environment management.

## Organizing Your Freesurfer Analyses

During Freesurfer installation, several environment variables are set in addition to `FREESURFER_HOME`.
One of these is `SUBJECTS_DIR`, which points to a directory containing the analysis output for all your subjects.

The `fs_subj_dir` function will return the path to this Freesurfer subjects directory if it's already set.
If not, it defaults to `file.path(fs_dir(), "subjects")`.

```{r fs_subj_dir_example}
fs_subj_dir()
```

You also have the flexibility to explicitly set the `freesurfer.subj_dir` option or `SUBJECTS_DIR` environment variable.

This default `SUBJECTS_DIR` setup in Freesurfer allows you to simply specify a subject identifier when running analyses, rather than needing to provide a specific file path or multiple intermediate files.

However, this default setup might not be ideal if you prefer to structure your data differently.
For example, if you have data from multiple studies, you might want to organize them into different folders located in various places.

Some Freesurfer functions rely on the `SUBJECTS_DIR` variable to run.
These functions typically take the subject name as their primary argument, rather than a file path, which is a more common approach in other tools.
To offer maximum flexibility, `freesurfer` allows most functions to specify a file or a different directory, rather than solely relying on the `SUBJECTS_DIR` and subject name.
Let's look at an example: the `asegstats2table` Freesurfer function.

Freesurfer performs segmentations of anatomical images into distinct structures, and it provides associated statistics for each region, such as volume and mean intensity.
The `asegstats2table` function transforms these anatomical segmentation statistics from images into a structured table.
By default, `asegstats2table` expects you to pass in a subject name rather than a direct file path.
However, the `freesurfer` `asegstats2table` function in R allows you to specify either the subject name *or* an alternative file name.

When you specify a file name, the function temporarily sets `SUBJECTS_DIR` to a temporary directory, copies your file there, executes the Freesurfer command, and then resets the `SUBJECTS_DIR` variable to its original state.
This provides a much more flexible workflow, allowing you to manage your data structure without overriding the default `SUBJECTS_DIR`.
This means you can have separate folders for subjects and easily read in data by simply switching the `subj_dir` argument in the R function.
The lookup order for using a subject directory is:

1. function argument  
1. `options(freesurfer.subj_dir = "/path/to/freesurfer/subjects")`  
1. System variable (`Sys.getenv("FREESURFER_SUBJ_DIR")`)
1. `file.path(fs_dir(), "subjects")`
    - Which _may_ utilise educated guesses based on common Freesurfer installation paths.

